Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-05 latex/main.tex
Add inference resources figure
Move robust recombs to methods

Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM (~5 CPU years; Figure SX).
The ARG contains 2,689,054 nodes representing the genealogical relationships
among 2,482,157 sampled genomes in the Viridian v04 dataset,
which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
Of the nodes, 855 represent recombination events,
of which 356 have a minimum of four loci supporting the non-focal parent on both sides of the breakpoint
(we refer to these as “robust” recombination events, Figure SX; see Document S1),
reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.

@@

@@

\section{Major deletions in the ARG}\label{sec_supplement_dels}
% Referenced by the main Results.
Basically, we want to say we did a lot of work to integrate signals of major
indels into the ARG, making it an even more powerful resource overall. Also,
indels are encoded by the Mutation-Annotated Tree format used by UShER
phylogenies (though PanMAT does this).

@@

\section{Identification of robust recombination
events}\label{sec_supplement_robust_recombs}
Before characterizing recombination events in detail, we sought to remove
probable artifacts. Systematic errors in SARS-CoV-2 sequences associated with
sequencing protocols have been documented (Turakhia et al., 2022), which can
confound recombination detection (De Maio et al., 2024). By inspecting copying
patterns at sites that differ between recombinant parents, we find a convincing
signal when treating adjacent and near-adjacent sites as a single locus.
Artifactual recombination events can be characterized as having fewer than four
such loci supporting the left or right parent on either side of a suggested
breakpoint (STAR Methods). Of the 929 recombination events in the ARG, this
measure identifies 543 as artifactual, leaving 386 “robust” events (Figure
S1A). The copying patterns depicted in Figure S1B illustrate the genomic
evidence for three recombination events identified as artifactual, compared to
a marginally robust one.

Evidence that our simple filter effectively flags artifacts can be seen in the
clear association between the AmpliSeq V1 kit and recombination events that we
identify as artifactual. This primer scheme was used to sequence XX of the
samples associated with an artifactual recombination event, in contrast to 15
of the “robust” events (Figure 2A), strongly implicating it as a common source
of problematic data. Through further inspection, we observed that many
artifactual recombination events contain stretches of missing data
characteristic of amplicon dropouts (REF).

@@

\section*{Supplementary Figures}
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{samples-per-day}
    \caption{TODO caption}\label{fig:samples_per_day}
\end{figure}

>>>>

Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM 
(~5 CPU years; Figure~\ref{fig:inference_resources}).
The ARG contains 2,689,054 nodes representing the genealogical relationships
among 2,482,157 sampled genomes in the Viridian v04 dataset,
which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
Of the nodes, 855 represent recombination events,
of which 356 have a minimum of four loci supporting the non-focal parent on both sides of the breakpoint
(we refer to these as ``robust'' recombination events,
Figure~\ref{fig:recombinant_qc}; see STAR Methods)
reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.

@@

\subsubsection*{Identification of robust recombination events}
[TODO update refs!]

Before characterizing recombination events in detail, we sought to remove
probable artifacts. Systematic errors in SARS-CoV-2 sequences associated with
sequencing protocols have been documented (Turakhia et al., 2022), which can
confound recombination detection (De Maio et al., 2024). By inspecting copying
patterns at sites that differ between recombinant parents, we find a convincing
signal when treating adjacent and near-adjacent sites as a single locus.
Artifactual recombination events can be characterized as having fewer than four
such loci supporting the left or right parent on either side of a suggested
breakpoint (STAR Methods). Of the 929 recombination events in the ARG, this
measure identifies 543 as artifactual, leaving 386 ``robust'' events (Figure
S1A). The copying patterns depicted in Figure S1B illustrate the genomic
evidence for three recombination events identified as artifactual, compared to
a marginally robust one.

Evidence that our simple filter effectively flags artifacts can be seen in the
clear association between the AmpliSeq V1 kit and recombination events that we
identify as artifactual. This primer scheme was used to sequence XX of the
samples associated with an artifactual recombination event, in contrast to 15
of the ``robust'' events (Figure 2A), strongly implicating it as a common source
of problematic data. Through further inspection, we observed that many
artifactual recombination events contain stretches of missing data
characteristic of amplicon dropouts (REF).

@@

\section{Major deletions in the ARG}\label{sec:supplement_dels}
% Referenced by the main Results.
Basically, we want to say we did a lot of work to integrate signals of major
indels into the ARG, making it an even more powerful resource overall. Also,
indels are encoded by the Mutation-Annotated Tree format used by UShER
phylogenies (though PanMAT does this).

@@

@@

\section*{Supplementary Figures}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{samples-per-day}
    \caption{TODO caption}\label{fig:samples_per_day}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{inference-resources}
    \caption{TODO caption}\label{fig:inference_resources}
\end{figure}
